A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-06-11 inst/manuscript/manuscript.Rmd
minor updates pre zenodo

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: St. Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches in a finite population spatial context.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated data and real data. The real data is from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest, and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between design-based inference and model-based inference is small when GRTS samples are used but large when SRS samples are used, suggesting that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the conterminous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the halved average squared differences (semivariance) of responses whose separation (distance) falls within a separation class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the empirical semivariance increases steadily with distance (meaning that observations near one another tend to be more similar than observations far apart from one another).

@@

95% interval coverage of both ZMMI and Hg ppb for all four sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}. When $n = 50$, interval coverage for both responses is too low, though interval coverage is higher for ZMMI (symmetric response) than for Hg ppb (skewed response). When $n = 100$, ZMMI interval coverage is approximately 95% except for GRTS-DB, which has coverage around 92%, while Hg ppb interval coverage ranges from approximately 90% (GRTS-DB) to 93% (GRTS-MB). When $n = 200$, ZMMI interval coverage is approximately 95% while Hg ppb interval coverage ranges from approximately 93% (GRTS-DB) to 95% (GRTS-MB).  As with the simulated data, coverages for the NLA data tend to increase with the sample sizes, coverages tend to be higher for symmetric responses than for skewed responses, and the local neighborhood variance was slightly too small for small $n$, yielding slightly lower interval coverages than the other sampling-inference combinations. Recall that model-based inference defines interval coverage properties across realized populations. With the simulated data, we evaluated interval coverage across realized populations, but for the NLA data, we evaluated interval coverage within a single realized population. We did find that model-based coverages were similar to the design-based coverages, however, suggesting that for some realized populations it is reasonable to heuristically view data from separate random samples as being from approximately separate realized populations. But generally, if model-based intervals constructed from many random samples of a single realized population show improper coverage, this does not necessarily imply a deficiency in model-based inference. Tables for interval coverage in all six simulation scenarios are provided in the supporting information.

@@

This manuscript has a supplementary \textbf{\textsf{R}} package that contains all of the data and code used in its creation. The supplementary \textbf{\textsf{R}} package is hosted on GitHub and archived on Zenodo. The Zenodo DOI is located here and in the references.

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: St. Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches in a finite population spatial context.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated data and real data. The real data is from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest, and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between design-based inference and model-based inference is small when GRTS samples are used but large when SRS samples are used, suggesting that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: references.bib
linenumbers: false
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the conterminous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Fig. \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Fig. \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the halved average squared differences (semivariance) of responses whose separation (distance) falls within a separation class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Fig. \ref{fig:zmmi}), as the empirical semivariance increases steadily with distance (meaning that observations near one another tend to be more similar than observations far apart from one another).

@@

95% interval coverage of both ZMMI and Hg ppb for all four sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}. When $n = 50$, interval coverage for both responses is too low, though interval coverage is higher for ZMMI (symmetric response) than for Hg ppb (skewed response). When $n = 100$, ZMMI interval coverage is approximately 95% except for GRTS-DB, which has coverage around 92%, while Hg ppb interval coverage ranges from approximately 90% (GRTS-DB) to 93% (GRTS-MB). When $n = 200$, ZMMI interval coverage is approximately 95% while Hg ppb interval coverage ranges from approximately 93% (GRTS-DB) to 95% (GRTS-MB).  As with the simulated data, coverages for the NLA data tend to increase with the sample sizes, coverages tend to be higher for symmetric responses than for skewed responses, and the local neighborhood variance was slightly too small for small $n$, yielding slightly lower interval coverages than the other sampling-inference combinations.

Recall that model-based inference defines interval coverage properties across realized populations. With the simulated data, we evaluated interval coverage across realized populations, but for the NLA data, we evaluated interval coverage within a single realized population. We did find that model-based coverages were similar to the design-based coverages, however, suggesting that for some realized populations it is reasonable to heuristically view data from separate random samples as being from approximately separate realized populations. But generally, if model-based intervals constructed from many random samples of a single realized population show improper coverage, this does not necessarily imply a deficiency in model-based inference. Tables for interval coverage in all six simulation scenarios are provided in the supporting information.

@@

This manuscript has a supplementary \textbf{\textsf{R}} package that contains all of the data, code, and figures used in its creation. All figures were created using the ggplot2 \textbf{\textsf{R}} package [@wickham2016ggplot2]. The supplementary \textbf{\textsf{R}} package is hosted on GitHub and archived on Zenodo. The Zenodo DOI is located here and in the references.

### 2022-06-11 inst/manuscript/manuscript.tex
minor updates pre zenodo

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less (be
more precise) than parameter estimates from samples lacking spatial
balance (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm to see widespread use was the generalized random tessellation
stratified (GRTS) algorithm (Stevens and Olsen, 2004). After the GRTS
algorithm was developed, several other spatially balanced sampling
algorithms emerged, including stratified sampling with compact
geographical strata (Walvoort et al., 2010), the local pivotal method
(Grafström et al., 2012; Grafström and Matei, 2018), spatially
correlated Poisson sampling (Grafström, 2012), balanced acceptance
sampling (Robertson et al., 2013), within-sample-distance sampling
(Benedetti and Piersimoni, 2017), and Halton iterative partitioning
sampling (Robertson et al., 2018). In this manuscript, we select
spatially balanced samples using the GRTS algorithm because it is
readily available in the \texttt{spsurvey} \textbf{\textsf{R}} package
(Dumelle et al., 2022) and naturally accommodates finite and infinite
sampling frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the
\texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle et al., 2022).
FPBK was implemented using the \texttt{sptotal} \textbf{\textsf{R}}
package (Higham et al., 2021a).

@@

We evaluated performance of the four sampling-inference combinations in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error (DRE). The three sample sizes (\(n\)) were \(n = 50, n = 100,\)
and \(n = 200\). Samples were always selected from a population size
(\(N\)) of \(N = 900\). The two location layouts were random and
gridded. Locations in the random layout were randomly generated inside
the unit square (\([0, 1] \times [0, 1]\)). Locations in the gridded
layout were placed on a fixed, equally spaced grid inside the unit
square. The two response types were normal and skewed. For the normal
response type, the response was simulated using mean-zero random errors
with the exponential covariance (Equation\(~\)\ref{equation:expcov}) for
three proportions of dependent random error (DRE): 0\% DRE, 50\% DRE,
and 90\% DRE. Recall the proportion of DRE is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the DRE variance and independent random error (IRE)
variance from Equation\(~\)\ref{equation:expcov}, respectively. The
total variance, \(\sigma^2_1 + \sigma^2_2\), was always 2. The distance
parameter was always \(\sqrt{2} / 3\), chosen so that the correlation in
the DRE decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the skewed
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
skewed random variable whose total variance was 2. The skewed responses
were used to evaluate performance of the sampling-inference approaches
for data that were not normally distributed but were still estimated
using REML, which relies on a normal log-likelihood. Figure
\ref{fig:sim_pops} shows an example of a realized population for the
normal and skewed responses using the random location layout and 50\%
DRE.

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within a separation class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the empirical
semivariance increases steadily with distance (meaning that observations
near one another tend to be more similar than observations far apart
from one another).

@@

95\% interval coverage of both ZMMI and Hg ppb for all four
sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}.
When \(n = 50\), interval coverage for both responses is too low, though
interval coverage is higher for ZMMI (symmetric response) than for Hg
ppb (skewed response). When \(n = 100\), ZMMI interval coverage is
approximately 95\% except for GRTS-DB, which has coverage around 92\%,
while Hg ppb interval coverage ranges from approximately 90\% (GRTS-DB)
to 93\% (GRTS-MB). When \(n = 200\), ZMMI interval coverage is
approximately 95\% while Hg ppb interval coverage ranges from
approximately 93\% (GRTS-DB) to 95\% (GRTS-MB). As with the simulated
data, coverages for the NLA data tend to increase with the sample sizes,
coverages tend to be higher for symmetric responses than for skewed
responses, and the local neighborhood variance was slightly too small
for small \(n\), yielding slightly lower interval coverages than the
other sampling-inference combinations. Recall that model-based inference
defines interval coverage properties across realized populations. With
the simulated data, we evaluated interval coverage across realized
populations, but for the NLA data, we evaluated interval coverage within
a single realized population. We did find that model-based coverages
were similar to the design-based coverages, however, suggesting that for
some realized populations it is reasonable to heuristically view data
from separate random samples as being from approximately separate
realized populations. But generally, if model-based intervals
constructed from many random samples of a single realized population
show improper coverage, this does not necessarily imply a deficiency in
model-based inference. Tables for interval coverage in all six
simulation scenarios are provided in the supporting information.

@@

This manuscript has a supplementary \textbf{\textsf{R}} package that
contains all of the data and code used in its creation. The
supplementary \textbf{\textsf{R}} package is hosted on GitHub and
archived on Zenodo. The Zenodo DOI is located here and in the
references.

@@

>>>>

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less (be
more precise) than parameter estimates from samples lacking spatial
balance (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm to see widespread use was the generalized random tessellation
stratified (GRTS) algorithm (Stevens and Olsen, 2004). After the GRTS
algorithm was developed, several other spatially balanced sampling
algorithms emerged, including stratified sampling with compact
geographical strata (Walvoort et al., 2010), the local pivotal method
(Grafström et al., 2012; Grafström and Matei, 2018), spatially
correlated Poisson sampling (Grafström, 2012), balanced acceptance
sampling (Robertson et al., 2013), within-sample-distance sampling
(Benedetti and Piersimoni, 2017), and Halton iterative partitioning
sampling (Robertson et al., 2018). In this manuscript, we select
spatially balanced samples using the GRTS algorithm because it is
readily available in the spsurvey \textbf{\textsf{R}} package (Dumelle
et al., 2022) and naturally accommodates finite and infinite sampling
frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the spsurvey
\textbf{\textsf{R}} package (Dumelle et al., 2022). FPBK was implemented
using the sptotal \textbf{\textsf{R}} package (Higham et al., 2021a).

@@

We evaluated performance of the four sampling-inference combinations in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error (DRE). The three sample sizes (\(n\)) were \(n = 50, n = 100,\)
and \(n = 200\). Samples were always selected from a population size
(\(N\)) of \(N = 900\). The two location layouts were random and
gridded. Locations in the random layout were randomly generated inside
the unit square (\([0, 1] \times [0, 1]\)). Locations in the gridded
layout were placed on a fixed, equally spaced grid inside the unit
square. The two response types were normal and skewed. For the normal
response type, the response was simulated using mean-zero random errors
with the exponential covariance (Equation\(~\)\ref{equation:expcov}) for
three proportions of dependent random error (DRE): 0\% DRE, 50\% DRE,
and 90\% DRE. Recall the proportion of DRE is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the DRE variance and independent random error (IRE)
variance from Equation\(~\)\ref{equation:expcov}, respectively. The
total variance, \(\sigma^2_1 + \sigma^2_2\), was always 2. The distance
parameter was always \(\sqrt{2} / 3\), chosen so that the correlation in
the DRE decayed to nearly zero at \(\sqrt{2}\), the largest possible
distance between two population units in the domain. For the skewed
response type, the response was first simulated using the same approach
as for the normal response type, except that the total variance was
0.6931 instead of 2. The response was then exponentiated, yielding a
skewed random variable whose total variance was 2. The skewed responses
were used to evaluate performance of the sampling-inference approaches
for data that were not normally distributed but were still estimated
using REML, which relies on a normal log-likelihood. Fig.
\ref{fig:sim_pops} shows an example of a realized population for the
normal and skewed responses using the random location layout and 50\%
DRE.

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Fig. \ref{fig:zmmi}. The ZMMI data tend to be highest near
the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Fig. \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within a separation class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Fig. \ref{fig:zmmi}), as the empirical
semivariance increases steadily with distance (meaning that observations
near one another tend to be more similar than observations far apart
from one another).

@@

95\% interval coverage of both ZMMI and Hg ppb for all four
sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}.
When \(n = 50\), interval coverage for both responses is too low, though
interval coverage is higher for ZMMI (symmetric response) than for Hg
ppb (skewed response). When \(n = 100\), ZMMI interval coverage is
approximately 95\% except for GRTS-DB, which has coverage around 92\%,
while Hg ppb interval coverage ranges from approximately 90\% (GRTS-DB)
to 93\% (GRTS-MB). When \(n = 200\), ZMMI interval coverage is
approximately 95\% while Hg ppb interval coverage ranges from
approximately 93\% (GRTS-DB) to 95\% (GRTS-MB). As with the simulated
data, coverages for the NLA data tend to increase with the sample sizes,
coverages tend to be higher for symmetric responses than for skewed
responses, and the local neighborhood variance was slightly too small
for small \(n\), yielding slightly lower interval coverages than the
other sampling-inference combinations.

Recall that model-based inference defines interval coverage properties
across realized populations. With the simulated data, we evaluated
interval coverage across realized populations, but for the NLA data, we
evaluated interval coverage within a single realized population. We did
find that model-based coverages were similar to the design-based
coverages, however, suggesting that for some realized populations it is
reasonable to heuristically view data from separate random samples as
being from approximately separate realized populations. But generally,
if model-based intervals constructed from many random samples of a
single realized population show improper coverage, this does not
necessarily imply a deficiency in model-based inference. Tables for
interval coverage in all six simulation scenarios are provided in the
supporting information.

@@

This manuscript has a supplementary \textbf{\textsf{R}} package that
contains all of the data, code, and figures used in its creation. All
figures were created using the ggplot2 \textbf{\textsf{R}} package
(Wickham, 2016). The supplementary \textbf{\textsf{R}} package is hosted
on GitHub and archived on Zenodo. The Zenodo DOI is located here and in
the references.

@@

\leavevmode\hypertarget{ref-wickham2016ggplot2}{}%
Wickham, H., 2016. ggplot2: Elegant graphics for data analysis.
Springer.
